Categories of Optics

### 2018-09-07 optics.tex
Couple more

When implementing optics, library authors often use a form known as the \emph{profunctor encoding}, which at first glance is completely different to that given in Section~\ref{sec:optics}. As this was being written, Milewski~\cite{ProfunctorOpticsPost} and Boisseau and Gibbons~\cite{YouNeeda} independently described the isomorphism between optics and their profunctor encoding. In Section~\ref{sec:profunctor-optics} we review this isomorphism and verify that the folklore profunctor optic laws are equivalent to lawfulness as defined here.

@@

  Suppose we have two lawful optics $\rep{l}{r} : R \hto S$ and $\rep{l'}{r'} : S \hto A$ with residuals $M$ and $N$ respectively. We must show that $\rep{l;(M\act l')}{(M\act r');r}$ is also a lawful optic. Showing the first law is straightforward:
  \begin{align*}
    \outside(\rep{l; (M\act l')}{(M\act r') ; r})
    &= l ; (M \act l') ; (M \act r') ; r \\
    &= l ; (M \act l'r') ; r \\
    &= l ; (M \act \id_{NA}) ; r \\
    &= l ; r \\
    &= \id_S
  \end{align*}

  For the second law, we must show that
  \[ \repthree{ l;(M\act l')}{(M\act r'); r;l;(M\act l')}{(M\act r') ; r} = \repthree{l;(M\act l')}{\id_{MNA}}{(M\act r') ; r}. \]
  The idea is that, by the lawfulness of $\rep{l}{r}$ and $\rep{l'}{r'}$, there are chains of coend relations that prove
  \begin{align*}
  \repthree{l}{r;l}{r} &= \repthree{l}{\id_{M \act S}}{r} \\
  \repthree{l'}{r';l'}{r'} &= \repthree{l'}{\id_{N \act A}}{r'}
  \end{align*}
  The result is achieved by splicing these chains of relations together in the following way.

@@

Now the type of lenses $(S, S') \hto (A, A')$ is the direct translation of the set of natural transformations $(U-)(A,A') \Rightarrow (U-)(S,S')$:
\begin{minted}{haskell}
type Lens s s' a a' = forall p. Strong p => p a a' -> p s s'
\end{minted}
where we use parametricity in \mintinline{haskell}{p} as a proxy for naturality. A profunctor lens 
\begin{minted}{haskell}
l :: forall p. Strong p => p a a -> p s s
\end{minted}
is lawful if it is monoidal as a natural transformation. In code this is:
\begin{minted}{haskell}
l id == id
l (Procompose p q) == Procompose (l p) (l q)
\end{minted}
where
\begin{minted}{haskell}
data Procompose p q d c where
  Procompose :: p x c -> q d x -> Procompose p q d c
\end{minted}
equipped with appropriate \mintinline{haskell}{Profunctor} and \mintinline{haskell}{Strong} instances denotes profunctor/Tambara module composition.

>>>>

When implementing optics, library authors often use a form known as the \emph{profunctor encoding}, which at first glance is completely different to that given in Section~\ref{sec:optics}. (The Haskell \lenslib{} library itself actually uses a variant called the \emph{van Laarhoven encoding}, for both historical and efficiency reasons.) As this was being written, Milewski~\cite{ProfunctorOpticsPost} and Boisseau and Gibbons~\cite{YouNeeda} independently described the isomorphism between optics and their profunctor encoding. In Section~\ref{sec:profunctor-optics} we review this isomorphism and verify that the folklore profunctor optic laws are equivalent to lawfulness as defined here.

@@

  Suppose we have two lawful optics $\rep{l}{r} : R \hto S$ and $\rep{l'}{r'} : S \hto A$ with residuals $M$ and $N$ respectively. We must show that $\rep{l;(M\act l')}{(M\act r');r}$ is also a lawful optic. Showing the first law is straightforward:
  \begin{align*}
    \outside(\rep{l; (M\act l')}{(M\act r') ; r})
    &= l ; (M \act l') ; (M \act r') ; r \\
    &= l ; (M \act l'r') ; r \\
    &= l ; (M \act \id_{N \act A}) ; r \\
    &= l ; r \\
    &= \id_S
  \end{align*}

  For the second law, we must show that
  \[ \repthree{ l;(M\act l')}{(M\act r'); r;l;(M\act l')}{(M\act r') ; r} = \repthree{l;(M\act l')}{\id_{M \act N \act A}}{(M\act r') ; r}. \]
  The idea is that, by the lawfulness of $\rep{l}{r}$ and $\rep{l'}{r'}$, there are chains of coend relations that prove
  \begin{align*}
  \repthree{l}{r;l}{r} &= \repthree{l}{\id_{M \act S}}{r} \\
  \repthree{l'}{r';l'}{r'} &= \repthree{l'}{\id_{N \act A}}{r'}
  \end{align*}
  The result is achieved by splicing these chains of relations together in the following way.

@@

Now the type of lenses $(S, S') \hto (A, A')$ is the direct translation of the set of natural transformations $(U-)(A,A') \Rightarrow (U-)(S,S')$:
\begin{minted}{haskell}
type Lens s s' a a' = forall p. Strong p => p a a' -> p s s'
\end{minted}
where we use parametricity in \mintinline{haskell}{p} as a proxy for naturality. A profunctor lens 
\begin{minted}{haskell}
l :: forall p. Strong p => p a a -> p s s
\end{minted}
is lawful if it is monoidal as a natural transformation. In code this is:
\begin{minted}{haskell}
l id == id
l (Procompose p q) == Procompose (l p) (l q)
\end{minted}
where
\begin{minted}{haskell}
data Procompose p q d c where
  Procompose :: p x c -> q d x -> Procompose p q d c
\end{minted}
denotes profunctor/Tambara module composition, once equipped with appropriate \mintinline{haskell}{Profunctor} and \mintinline{haskell}{Strong} instances.
